---
title: "Introduction to Relational Database Management Systems"
date: "2023/7/22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Creating and Manipulating Databases in R Using `DBI`

## Connecting to Databases

`DBI` is a library which is included in the package `RSQLite`. The package could be installed by the following code:

```{r eval=FALSE}
# the code is not executed as the package has already been installed
install.packages("RSQLite")
```

And then we load the required library `DBI`:
```{r}
# load the required library
library(DBI)
```

We then use the function `dbConnect()` to create an object, `conn`, to connect to the SQLite driver to manipulate the database `university.db`. Before that, we need to check whether the database `university.db` exists. If the database `university.db` exists, the following code chunk will remove it.

```{r results='hide'}
# delete "university.db" if already exists
if (file.exists("university.db"))
  file.remove("university.db")
```

We use `dbConnect()` to connect to the database:
```{r}
conn <- dbConnect(RSQLite::SQLite(), "university.db")
```

We can list all tables in `university.db` using the function `dbListTable()` from `DBI`.

```{r}
# list all table
dbListTables(conn)
```

Nothing is returned because we have not created any tables in the database yet.

## Creating Tables

We are going to create some tables to the database `university.db`. We will create the tables using data saved in the CSV files. We first read the CSV files into `data.frame` in R:

```{r}
# read the CSV files into data.frame in R
course <- read.csv("Course Files/Block 3/course.csv", header = TRUE)
student <- read.csv("Course Files/Block 3/student.csv", header = TRUE)
grade <- read.csv("Course Files/Block 3/grade.csv", header = TRUE)
```

We then copy the data frames `student`, `grade` and `course` to tables in the database `university.db` using `DBI`’s `dbWriteTable()` function:

```{r}
dbWriteTable(conn, "Course", course)
dbWriteTable(conn, "Student", student)
dbWriteTable(conn, "Grade", grade)
```

Now we can see there are three tables in the database:

```{r}
# list all tables
dbListTables(conn)
```

We can also browse any table in the database using the function `dbReadTable()` from `DBI`.

```{r}
# browse the table
dbReadTable(conn, "Student")
```

Or we can see the attributes of a table (e.g. `Student`) by the function `dbListFields()`:

```{r}
# browse the attributes of a table
dbListFields(conn, "Student")
```

## Manipulating Databases

The simplest way to manipulate databases is to use the `dbExecute()` function. This function executes SQL statements and returns the number of rows affected.

### Adding a New Table

We can add a new table by using the function `dbCreateTable()`. Alternatively, you can use `dbExecute()` to run the SQL command to create a new table.

```{r}
# add a new table using dbCreateTable() function
dbCreateTable(conn, "Teacher", c(staff_id = "TEXT", name = "TEXT"))
```

Alternatively

```{r results='hide'}
# add a new table using dbExecute() function
dbExecute(conn,
          "CREATE TABLE Other_Staff (
          staff_id TEXT PRIMARY KEY,
          name TEXT)")
```

List of tables after adding:

```{r}
# list all tables
dbListTables(conn)
```

### Deleting a Table

We can remove a table by using the function `dbRemoveTable()`:

```{r}
# delete a table using dbRemoveTable() function
dbRemoveTable(conn, "Teacher")
```

Alternatively, we can use `dbExecute()` function:

```{r results='hide'}
# delete a table using dbExecute() function
dbExecute(conn,
          "DROP TABLE Other_Staff")
```

When we list the tables, we can now see three tables.

```{r}
# list all tables
dbListTables(conn)
```

### Inserting Tuples/Rows

Below we insert the year 1 student “Harper Taylor” with student ID 202029744 to Student by using the function `dbAppendTable()`:

```{r results='hide'}
# insert tuples/rows using dbAppendTable() function
dbAppendTable(conn, "Student", data.frame(student_id = "202029744",
                                          name = "Harper Taylor",
                                          year = 1))
```

Alternatively, we can use `dbExecute()` function:

```{r results='hide'}
# insert tuples/rows using dbExecute() function
dbExecute(conn,
          "INSERT INTO Student VALUES(19890604, 'Never Forget', 1)")
```

When we browse the table, we can see the new row has been added.

```{r}
# browse the table
dbReadTable(conn, "Student")
```

### Updating Tuples/Rows

Below, we update the student ID of student Harper Taylor to 201929744 by `dbExecute()`. There is no specific function for updating a row in `DBI`.

```{r results='hide'}
# update tuples/rows
dbExecute(conn, "Update Student SET student_id = '201929744'
                                WHERE name = 'Harper Taylor'")
```

When we browse the table, we can see the new row has been added.

```{r}
# browse the table
dbReadTable(conn, "Student")
```


### Deleting Tuples/Rows

Below, we delete the record for the student Harper Taylor and Never Forget from table `Student` using `dbExecute()`. There is no specific function for deleting a row in `DBI`.

```{r}
dbExecute(conn,
          "DELETE FROM Student WHERE name = 'Harper Taylor' OR name = 'Never Forget'")
```

When we browse the table, we can see the rows have been removed.

```{r}
# browse the table
dbReadTable(conn, "Student")
```

## Disconnecting From the Database

After we finish manipulating the database, we can close the connection using the function `dbDisconnect()` from `DBI`:

```{r}
# disconnecting from the database
dbDisconnect(conn)
```

## Querying Databases Using `DBI`

We have just disconnected from the database. We need to reconnect to the database by the following codes in order to perform querying databases.

```{r}
# re-connect to the database
conn <- dbConnect(RSQLite::SQLite(), "university.db")
dbListTables(conn)
```

Once we formulate the query into an `SQL SELECT` statement, we can get the query result in `R` using the function `dbGetQuery()` or `dbSendQuery()` from `DBI`. The following examples show how we can run queries from within `R`.

### Getting Grades of a Course With `course_id` "ST101"

```{r}
# getting grades of a course using dbGetQuery()
q1 <- dbGetQuery(conn, "SELECT final_mark
                        FROM Grade
                        WHERE course_id = 'ST101'")
q1
```

The second argument in `dbGetQuery()` is the `SQL` query statement sent using `DB Browser` for `SQLite`.
<br>

Note that the `dbGetQuery()` returns a `data.frame`.

```{r}
# data type of q1
class(q1)
```

Alternatively, we can use `dbSendQuery()` and `dbFetch()`:

```{r}
# getting grades of a course using dbSendQuery()
q1 <- dbSendQuery(conn, "SELECT final_mark
                         FROM Grade
                         WHERE course_id = 'ST101'")
q1
```

Note `dbSendQuery()` only sends and executes the `SQL` query to the database engine. It does not extract any records. When we run `dbFetch()`, the executed query result will then be fetched.

```{r}
# fetch the result
dbFetch(q1)
```

### Getting Names of Students in Alphabetical Order

```{r}
# getting names of students in alphabetical order
dbGetQuery(conn, "SELECT Student.name
                  FROM Grade, Student
                  WHERE Grade.course_id = 'ST101' AND student.student_id = Grade.student_id
                  ORDER BY Student.name")
```

Or we can do it with `NATURAL JOIN`:

```{r}
# using NATURAL JOIN
dbGetQuery(conn, "SELECT Student.name
                  FROM Student NATURAL JOIN Grade
                  WHERE course_id = 'ST101'
                  ORDER BY Student.name")
```

### Getting Courses Taken by Students Ava Smith or Freddie Harris

```{r}
# getting courses taken by students
dbGetQuery(conn, "SELECT DISTINCT Course.name
                  FROM Student, Grade, Course
                  WHERE (Student.name = 'Ava Smith' OR Student.name = 'Freddie Harris')
                        AND Student.student_id = Grade.student_id
                        AND Course.course_id = Grade.course_id")
```

Or we can do it using `JOIN`:

```{r}
# using JOIN
dbGetQuery(conn, "SELECT DISTINCT Course.name
                  FROM (student NATURAL JOIN Grade) S JOIN Course on Course.course_id = S.course_id
                  WHERE S.name = 'Ava Smith' OR S.name = 'Freddie Harris'")

```


# Useful Links and Resources

* [Using **DBI**](https://solutions.posit.co/connections/db/r-packages/dbi/)
* [**DBI** reference manual](chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://cran.r-project.org/web/packages/DBI/DBI.pdf)
* [**RSQLite** vignettes](https://cran.r-project.org/web/packages/RSQLite/vignettes/RSQLite.html)