---
title: "Creating, Manipulating and Querying Databases in `R` Using `DBI` and `dplyr`"
date: "`r Sys.Date()`"
author: Steven Wun
output:
  pdf_document: default
  rmdformats::material:
    highlight: kate
  word_document: default
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Interacting with the SQLite driver in `R`

`DBI` is the library used to interact with the SQLite driver and perform various operations in manipulating the database. The library `DBI` is included in the package `RSQLite`. The package could be installed by the following code:

```{r eval=FALSE}
# the code is not executed as the package has already been installed
install.packages("RSQLite")
```

We can load the required library `DBI` by the following code:
```{r}
# load the required library
library(DBI)
```

## Connecting to Databases

Thereafter, the first step is to connect to the database. We use the function `dbConnect()` to create an object, `conn`, to connect to the SQLite driver to manipulate the database `university.db`. Before that, we need to check whether the database `university.db` exists. If the database `university.db` exists, the following code chunk will remove it.

```{r results='hide'}
# delete "university.db" if already exists
if (file.exists("university.db"))
  file.remove("university.db")
```

We use `dbConnect()` to connect to the database:
```{r}
conn <- dbConnect(RSQLite::SQLite(), "university.db")
```

We have created the database `university.db`. However, it is empty now. We can list all tables in `university.db` using the function `dbListTable()` from `DBI`.

```{r}
# list all table
dbListTables(conn)
```

Nothing is returned because we have not created any tables in the database yet.

## Creating Tables

We are going to create some tables to the database `university.db`. We will create the tables using data saved in the CSV files. We first read the CSV files into `data.frame` in R:

```{r}
# assign the CSV files on GitHub site to objects 
course.csv <- "https://raw.githubusercontent.com/stevenkhwun/ST2195/main/Course%20Files/Block%203/course.csv"
student.csv <- "https://raw.githubusercontent.com/stevenkhwun/ST2195/main/Course%20Files/Block%203/student.csv"
grade.csv <- "https://raw.githubusercontent.com/stevenkhwun/ST2195/main/Course%20Files/Block%203/grade.csv"

# read the CSV files into data.frame in R
course <- read.csv(course.csv, header = TRUE)
student <- read.csv(student.csv, header = TRUE)
grade <- read.csv(grade.csv, header = TRUE)
```

We then copy the data frames `student`, `grade` and `course` to tables in the database `university.db` using `DBI`’s `dbWriteTable()` function:

```{r}
dbWriteTable(conn, "Course", course)
dbWriteTable(conn, "Student", student)
dbWriteTable(conn, "Grade", grade)
```

Now we can see there are three tables in the database:

```{r}
# list all tables
dbListTables(conn)
```

We can also browse any table in the database using the function `dbReadTable()` from `DBI`.

```{r}
# browse the table
dbReadTable(conn, "Student")
```

Or we can see the attributes of a table (e.g. `Student`) by the function `dbListFields()`:

```{r}
# browse the attributes of a table
dbListFields(conn, "Student")
```

## Disconnecting From the Database

After we finish creating the database, we can close the connection using the function `dbDisconnect()` from `DBI`:

```{r}
# disconnecting from the database
dbDisconnect(conn)
```


# Manipulating Databases

The simplest way to manipulate databases is to use the `dbExecute()` function. This function executes SQL statements and returns the number of rows affected.
<br>

We have just disconnected from the database. We need to reconnect to the database by the following codes in order to perform manipulating the databases.

```{r}
# re-connect to the database
conn <- dbConnect(RSQLite::SQLite(), "university.db")
dbListTables(conn)
```


## Adding a New Table

We can add a new table by using the function `dbCreateTable()`. Alternatively, you can use `dbExecute()` to run the SQL command to create a new table.

```{r}
# add a new table using dbCreateTable() function
dbCreateTable(conn, "Teacher", c(staff_id = "TEXT", name = "TEXT"))
```

Alternatively

```{r results='hide'}
# add a new table using dbExecute() function
dbExecute(conn,
          "CREATE TABLE Other_Staff (
          staff_id TEXT PRIMARY KEY,
          name TEXT)")
```

List of tables after adding:

```{r}
# list all tables
dbListTables(conn)
```

## Deleting a Table

We can remove a table by using the function `dbRemoveTable()`:

```{r}
# delete a table using dbRemoveTable() function
dbRemoveTable(conn, "Teacher")
```

Alternatively, we can use `dbExecute()` function:

```{r results='hide'}
# delete a table using dbExecute() function
dbExecute(conn,
          "DROP TABLE Other_Staff")
```

When we list the tables, we can now see three tables.

```{r}
# list all tables
dbListTables(conn)
```

## Inserting Tuples/Rows

Below we insert the year 1 student “Harper Taylor” with student ID 202029744 to Student by using the function `dbAppendTable()`:

```{r results='hide'}
# insert tuples/rows using dbAppendTable() function
dbAppendTable(conn, "Student", data.frame(student_id = "202029744",
                                          name = "Harper Taylor",
                                          year = 1))
```

Alternatively, we can use `dbExecute()` function:

```{r results='hide'}
# insert tuples/rows using dbExecute() function
dbExecute(conn,
          "INSERT INTO Student VALUES(201989604, 'Paul Simon', 1)")
```

When we browse the table, we can see the new row has been added.

```{r}
# browse the table
dbReadTable(conn, "Student")
```

## Updating Tuples/Rows

Below, we update the student ID of student Harper Taylor to 201929744 by `dbExecute()`. There is no specific function for updating a row in `DBI`.

```{r results='hide'}
# update tuples/rows
dbExecute(conn, "UPDATE Student SET student_id = '201929744'
                                WHERE name = 'Harper Taylor'")
```

When we browse the table, we can see the new row has been added.

```{r}
# browse the table
dbReadTable(conn, "Student")
```


## Deleting Tuples/Rows

Below, we delete the record for the student Harper Taylor and Never Forget from table `Student` using `dbExecute()`. There is no specific function for deleting a row in `DBI`.

```{r}
dbExecute(conn,
          "DELETE FROM Student WHERE name = 'Harper Taylor' OR name = 'Paul Simon'")
```

When we browse the table, we can see the rows have been removed.

```{r}
# browse the table
dbReadTable(conn, "Student")
```


# Basic SQL/SQLite Syntax and Queries


Once we formulate the query into an `SQL SELECT` statement, we can get the query result in `R` using the function `dbGetQuery()` or `dbSendQuery()` from `DBI`. The following examples show how we can run queries from within `R`.

## Conditions: Basics
Getting Grades of a Course With `course_id` "ST101"

```{r}
# getting grades of a course using dbGetQuery()
q1 <- dbGetQuery(conn, "SELECT final_mark
                        FROM Grade
                        WHERE course_id = 'ST101'")
q1
```

The second argument in `dbGetQuery()` is the `SQL` query statement sent using `DB Browser` for `SQLite`.
<br>

Note that the `dbGetQuery()` returns a `data.frame`.

```{r}
# data type of q1
class(q1)
```

Alternatively, we can use `dbSendQuery()` and `dbFetch()`:

```{r}
# getting grades of a course using dbSendQuery()
q1 <- dbSendQuery(conn, "SELECT final_mark
                         FROM Grade
                         WHERE course_id = 'ST101'")
q1
```

Note `dbSendQuery()` only sends and executes the `SQL` query to the database engine. It does not extract any records. When we run `dbFetch()`, the executed query result will then be fetched.

```{r}
# fetch the result
dbFetch(q1)
```

## Conditions: The operator `*`
We would like to return all attributes of the students who are in ST101:

```{r}
# getting all the attributes of the students who are in sT101
dbGetQuery(conn, "SELECT *
                  FROM Grade
                  WHERE course_id = 'ST101'")
```

## Conditions: `iif()` function
In SQLite, `iif()` is a conditional function that returns the second or third argument based on the evaluation of the first argument.
<br>

It’s logically equivalent to `CASE WHEN X THEN Y ELSE Z END`.
<br>

`iif()` is an abbreviation for `Immediate IF`.

```{r}
# using iif()
dbGetQuery(conn, "SELECT *,
                  iif( final_mark > 60, 'A', 'F') AS LetterGrade
                  FROM Grade")
```

Refer to this [link](https://database.guide/how-iif-works-in-sqlite/) for more information.

## Conditions: `CASE` statement

The SQLite `CASE` expression evaluates a list of conditions and returns an expression based on the result of the evaluation.

The `CASE` expression is similar to the `IF-THEN-ELSE` statement in other programming languages.

```{r}
# using iif()
dbGetQuery(conn, "SELECT *,
                  CASE
                    WHEN final_mark >= 80 THEN 'A'
                    WHEN final_mark < 80 AND final_mark >= 60 THEN 'B'
                    ELSE 'C'
                    END LetterGrade
                  FROM Grade")
```

Refer to this [link](https://www.sqlitetutorial.net/sqlite-case/) for more information.


## Several Tables
Getting Names of Students in Alphabetical Order
<br>

Note the student name information is in the `Student` table whereas the information about which course the student took is in `Grade`. In order to perform the query we need to combine information from the `Student` and `Grade` tables.

```{r}
# getting names of students in alphabetical order
dbGetQuery(conn, "SELECT name
                  FROM Grade, Student
                  WHERE course_id = 'ST101' AND student.student_id = Grade.student_id
                  ORDER BY name")
```

Note that we don't need to specify the table name for the attributes `name` and `course_id` because attribute `name` is only in the table `Student` and the attribute `course_id` is only in the table `Grade`.

Or we can do it with `NATURAL JOIN`:

```{r}
# using NATURAL JOIN
dbGetQuery(conn, "SELECT Student.name
                  FROM Student NATURAL JOIN Grade
                  WHERE course_id = 'ST101'
                  ORDER BY Student.name")
```

## Multiple Conditions 
Getting Courses Taken by Students Ava Smith or Freddie Harris

```{r}
# getting courses taken by students
dbGetQuery(conn, "SELECT Course.name
                  FROM Student, Grade, Course
                  WHERE (Student.name = 'Ava Smith' OR Student.name = 'Freddie Harris')
                        AND Student.student_id = Grade.student_id
                        AND Course.course_id = Grade.course_id")
```

There are a few duplicate rows in the output, which we can remove using `DISTINCT`:

```{r}
# getting courses taken by students (duplicates removed)
dbGetQuery(conn, "SELECT DISTINCT Course.name
                  FROM Student, Grade, Course
                  WHERE (Student.name = 'Ava Smith' OR Student.name = 'Freddie Harris')
                        AND Student.student_id = Grade.student_id
                        AND Course.course_id = Grade.course_id")
```

Or we can do it using `JOIN`:

```{r}
# using JOIN
dbGetQuery(conn, "SELECT DISTINCT Course.name
                  FROM (student NATURAL JOIN Grade) S JOIN Course on Course.course_id = S.course_id
                  WHERE S.name = 'Ava Smith' OR S.name = 'Freddie Harris'")

```

## Aggregation and `GROUP BY`
We would like to calculate the average mark for each course according to the value of `course_id`:

```{r}
# calculate the average mark
dbGetQuery(conn, "SELECT course_id, AVG(final_mark)
                  FROM Grade
                  GROUP BY course_id")
```

The attribute name for the average mark looks different from other attributes. We can rename it using the `AS` clause:

```{r}
# calculate the average mark
dbGetQuery(conn, "SELECT course_id, AVG(final_mark) AS avg_mark
                  FROM Grade
                  GROUP BY course_id")
```

# SQL Joins

A `JOIN` clause combines rows from two or more tables based on related column(s) between them. The SQL language offers many different types of joins.

*   `INNER JOIN`: Select rows that have matching values in both tables based on the given columns
*   `NATURAL JOIN`: Similar to `INNER JOIN` except that there is no need to specify which columns are used for matching values
*   `OUTER JOIN`: Unlike `INNER JOIN`, unmatched rows in one or both tables can be returned. There are `LEFT`, `RIGHT` and `FULL OUTER JOIN`. `SQLite` only supports `LEFT OUTER JOIN`. For `LEFT OUTER JOIN`, all the records from the left table are included in the result.
*   `CROSS JOIN`: Return the Cartesian product of the two joined tables, by matching all the values from the left table with all the values from the right table.


## `INNER JOIN`

Below we get the records about students who took the course with course ID ST101 and sort the student names in alphabetical order. Without using Joins, we can implement the following codes:

```{r}
# without using join
dbGetQuery(conn, "SELECT *
                  FROM Student, Grade
                  WHERE Grade.course_id = 'ST101' AND Student.student_id = Grade.student_id
                  ORDER BY Student.name")
```

We can rewrite the above query using `INNER JOIN`:

```{r}
# using INNER JOIN
dbGetQuery(conn, "SELECT *
                  FROM Student JOIN Grade ON Student.student_id = Grade.student_id
                  WHERE course_id = 'ST101'
                  ORDER BY Student.name")
```

Note that:

*   We write `JOIN` instead of `INNER JOIN` as by default `INNER JOIN` is used when you do not specify the join type.
*   The `ON` keyword specifies on what condition you want to join the tables.
*   If you look at the result, `student_id` appears twice - this is because all the columns from both tables are returned.
*   By using the `JOIN` clause, we separate the logic of combining the tables (`Student.student_id = Grade.student_id`) and the other condition (`course_id = 'ST101'`), which makes the SQL query more readable.

Instead of joining using `ON`, we can use `USING` with `JOIN` if the columns that we are joining have the same name. For example:

```{r}
# 'JOIN' clause using 'USING'
dbGetQuery(conn, "SELECT *
                  FROM Student JOIN Grade USING (student_id)
                  WHERE course_id = 'ST101'
                  ORDER BY Student.name")
```

Note that:

*   The `USING` keyword specifies which column is used to select rows that have matching values in both tables.
*   If you look at the result, `student_id` appears only once now.

## `NATURAL JOIN`

```{r}
# using 'NATURAL JOIN'
dbGetQuery(conn, "SELECT *
                  FROM Student NATURAL JOIN Grade
                  WHERE course_id = 'ST101'
                  ORDER BY Student.name")
```

Note that:

*   We do not specify how to join the two tables. The join condition is automatically identified.
*   If you look at the result, `student_id` appears only once.

## `LEFT JOIN`

When we run the following `SQL` commands to use `INNER JOIN` to combine the tables `Student` and `Grade`

```{r}
# use 'INNER JOIN' to join tables
dbGetQuery(conn, "SELECT *
                  FROM Student INNER JOIN Grade USING (student_id)
                  ORDER BY Student.name")
```

the record of students Ben Johnson and Dan Norris are not shown, because there are not corresponding records for these two students in the table `Grade`. If we instead use `LEFT OUTER JOIN` to combine the tables `Student` and `Grade`:

```{r}
# use 'LEFT OUTER JOIN' to join tables
dbGetQuery(conn, "SELECT *
                  FROM Student LEFT JOIN Grade USING (student_id)
                  ORDER BY Student.name")
```

we get a result where:

*   All the students from the left table `Student` are included (including Ben Johnson and Dan Norris)
*   The students with no corresponding record in the right table `Grade`, have `NULL` value in attributes `course_id` and `final_mark` from `Grade`.


## `CROSS JOIN`

When we use `CROSS JOIN`:

```{r}
# use 'CROSS JOIN'
dbGetQuery(conn, "SELECT *
                  FROM Student CROSS JOIN Grade
                  ORDER BY Student.name")
```

we get 56 rows, which is number of rows in `Student` (7) times number of rows in `Grade` (8).


# Querying Databases in `R` Using `dplyr`

## Introduction to dplyr

`dplyr` is an `R` package for data manipulation. It provides a set of functions, named as verbs, that can be used to carry out a wide range of data manipulation operations:

*   The `mutate()` function adds new variables that are transformations of existing variables.
*   The `select()` function picks variables based on their names.
*   The `filter()` function picks cases based on their values.
*   The `summarize()` function reduces multiple values down to a single summary.
*   The `arrange()` function changes the ordering of the rows.

`dplyr` verbs operate on data frames, but they also, almost seamlessly apply to database tables! In particular `dplyr` allows you to use database tables as if they are data frames, by internally converting `dplyr` code into `SQL` commands using `dbplyr`. The following table compares the syntax used in `SQL` with `dplyr` syntax:

| Action         | `SQL`                                          | `dplyr`  |  
| -------------- | ---------------------------------------------- | ---------|
| select a column| SELECT                                         | select   |
| select a row   | WHERE                                          | filter   |
| sort           | ORDER BY                                       | arrange  |
| group          | GROUP BY                                       | group_by |
| aggregation    | aggregation functions (e.g. `AVG()`) in SELECT | summarize|

See the [SQL Transalation](https://dbplyr.tidyverse.org/articles/sql-translation.html) article in `dbplyr`'s pages for more information.

## Pipe Operator

All of the `dplyr` functions take a data frame (or a [`tibble`](https://tibble.tidyverse.org/)) as the first argument. In this way, the `%>%` operator from the `magrittr` `R` package can be used, so that user does not need to save intermediate objects or nest verbs. For example, the statement `x %>% f(y)` is equivalent to `f(x, y)`, and the result from one step is “piped” into the next step. You can think of the pipe operator as “then.”

## Querying Databases Using `dplyr`

### Connecting to the Database

We should have connected to the database `university.db`. If not, connect to the database by the method as mentioned above. In the meantime, let us see all the three tables `student`, `course` and `grade` in the database `university.db`.

```{r}
# list all the tables in university.db
dbListTables(conn)
```

### Creating a Reference to Table

```{r message=FALSE}
# creating a reference to table
library(dplyr)
student_db <- tbl(conn, "Student")
grade_db <- tbl(conn, "Grade")
course_db <- tbl(conn, "Course")
```

By creating references to the tables as done above, we can treat `student_db`, `grade_db`, and `course_db` as data frames, and use `dplyr` functionality to query the database.

### Getting Grades of the Course using `filter()`

```{r}
# getting grades of the course using filter()
q1 <- grade_db %>% filter(course_id == "ST101")
q1
```

The function `filter()` selects the rows in the `grade_db` which satisfy the condition `course_id == "ST101"`.
<br>

We can use the function `show_query()` to show the `SQL` query that `dbplyr` produced, when we run the code above:

```{r}
# show the SQL query produced
show_query(q1)
```

### Getting Names of Stuents in Alphabetic Order

If we want to work on more than one table in `dplyr`, we can use join or set operations. In this example we show how to use `inner_join()`. `arrange()` is then used to order the query result.

```{r message=FALSE}
# using 'inner_join()' and 'arrange()'
q2 <- inner_join(student_db, grade_db) %>% 
  filter(course_id == "ST101") %>% 
  select(name) %>% 
  arrange(name)
q2
```

The corresponding `SQL` query is:

```{r}
# show the SQL query produced
show_query(q2)
```

### Getting Courses Taken by Ava Smith or Freddie Harris

Here we use `inner_join` to specify which attribute should be used to join by, in this case `course_id`. As both `student_db` and the `course_db` have the attribute `name`, we use the argument `suffix` to rename the attribute `name` to `name.student` and `name.course`, correspondingly. In this way we eliminate ambiguity.

```{r message=FALSE}
# use the argument 'suffix'
q3 <- inner_join(student_db, grade_db, by = "student_id") %>% 
  inner_join(course_db, by = "course_id", suffix = c(".student", ".course")) %>% 
  filter(name.student == 'Ava Smith' | name.student == 'Freddie Harris') %>% 
  select(name.course) %>% 
  distinct()
q3
```

The corresponding `SQL` query is:

```{r}
# show the SQL query produced
show_query(q3)
```

Note the computer-generated `SQL` code that `dplyr` created internally is more complicated than the `SQL` code we wrote for the same query.

### Calculating Average Mark for Each Course

The combination of the verbs `group_by()` and `summarize()` are used to calculate the average `final_mark` for each `course_id`.

```{r message=FALSE}
# use of 'group_by()' and 'summarize()'
q4 <- grade_db %>% 
  group_by(course_id) %>% 
  summarize(avg_mark = mean(final_mark, na.rm = TRUE))
q4
```


The corresponding `SQL` query is:

```{r}
# show the SQL query produced
show_query(q4)
```

### Disconnecting From the Database

After we finish manipulating the database, we can close the connection using the function `dbDisconnect()` from `DBI`:

```{r}
# disconnect from the database
dbDisconnect(conn)
```

# Useful Links and Resources

* `DBI`
  * [Using **DBI**](https://solutions.posit.co/connections/db/r-packages/dbi/)
  * [**DBI** reference manual](chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://cran.r-project.org/web/packages/DBI/DBI.pdf)
  * [**RSQLite** vignettes](https://cran.r-project.org/web/packages/RSQLite/vignettes/RSQLite.html)

* `SQL` Joins
  * [SQLite join](https://www.sqlitetutorial.net/sqlite-join/) to learn more about how to join tables via SQLite
  * [SQLite select](https://www.sqlitetutorial.net/sqlite-select/) to learn more about how to query via SQLite
  * [SQL](https://en.wikipedia.org/wiki/SQL) from Wikipedia
  
* `dplyr`
  * [Using `dplyr` with databases](https://solutions.posit.co/connections/db/r-packages/dplyr/): A guide on how to use `dplyr` with database from RStudio
  * [`dplyr` vignettes](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html): A introduction to `dplyr`
  * [`dbplyr` SQL Translation](https://dbplyr.tidyverse.org/articles/sql-translation.html)


