---
title: "Data Wrangling Operations in `Python`"
date: "`r Sys.Date()`"
author: Steven Wun
output:
  rmdformats::material:
    highlight: kate
  pdf_document: default
  word_document: default
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

In order to run `Python` in `R`, we need to load the `reticulate` package first. Further, unless marked as `R code`, all the codes below are Python codes.

```{r warning=FALSE}
# R code
# load the required library to run Python in R
library(reticulate)
```

# Using Databases With `Python`

We import the module `sqlite3` and use the function `connect()` to create an object, `conn`, to connect to the `SQLite` driver to manipulate the database `University.db`. If the database `University.db` exists in your working directory, the following code chunk will remove it.


```{python}
# this makes sure you can run this notebook multiple times without errors
import os
try:
  os.remove('university.db')
except OSError:
  pass
```

```{python}
# connect to the SQLite driver
import sqlite3
conn = sqlite3.connect('university.db')
```

## Creating Tables Using `Python`

Now we are going to create some tables to the database `University.db`. Like before, we will create the tables using the data saved in the CSV files. We first load the CSV files into `DataFrame` in Python:

```{python}
# creating tables using python
import pandas as pd
student = pd.read_csv("Course Files/Block 3/student.csv")
course = pd.read_csv("Course Files/Block 3/course.csv")
grade = pd.read_csv("Course Files/Block 3/grade.csv")
```

We then write record stored in `DataFrame`s `student`, `grade` and `course` as tables to the database `University.db` using the `DataFrame` method `to_sql()`.

```{python results='hide'}
# write tables to the database
# 'index = False` to ensure the DataFrame row index is not written into the SQL tables
student.to_sql('Student', con = conn, index = False)
course.to_sql('Course', con = conn, index = False)
grade.to_sql('Grade', con = conn, index = False)
```

## Manipulate Databases Using `Python`

We can manipulate databases in `Python` by the `execute()` and `fetchall()` methods from the `sqlite3` module which performs `SQL` commands. This allows us to leverage the `SQL` commands we have learned to manipulate the databases in `Python`. We first need to create a cursor object `c`:

```{python}
# create a cursor object 'c'
c = conn.cursor()
```

After that we can execute the `SQL` commands we learned before using the function `execute()` and `fetchall()`. For example, if we want to get all the tables in the database, we can run:

```{python}
# get all the tables in a database
c.execute('''
SELECT name
FROM sqlite_master
WHERE type='table'
''')
```

The result is not returned until we run `fetchall()`:

```{python}
# return the result
c.fetchall()
```

We can see there are three tables in the database. If we want to browse the table `Student` we can run (here we display the results as `pandas` `DataFrame`):

```{python}
# browse the contents of a table
q = c.execute("SELECT * FROM Student").fetchall()
pd.DataFrame(q)
```

Note here we combine the use of `execute()` and `fetchall()` in one line.

### Add a New Table

We can add a new table by running the SQL command through `execute()`:

```{python}
# add a new table
c.execute('''
CREATE TABLE Teacher (staff_id TEXT PRIMARY KEY, name TEXT)
''')
conn.commit()     # save (commit) the changes
```

When we list the tables, we can see four tables.

```{python}
# list the tables
c.execute('''
SELECT name
FROM sqlite_master
WHERE type='table'
''').fetchall()
```

### Delete a Table

We can delete a table by running the `SQL` command through `execute()`:

```{python}
# delete a table
c.execute("DROP TABLE Teacher")
conn.commit()
```

When we list the tables, we can see three tables.

```{python}
# list the tables
c.execute('''
SELECT name
FROM sqlite_master
WHERE type='table'
''').fetchall()
```

### Insert Tuples/Rows

Insert the year 1 student Harper Taylor with student ID 202029744 to Student:

```{python}
# insert rows
c.execute("INSERT INTO Student VALUES(202029744, 'Harper Taylor', 1)")
conn.commit()
```

When we browse the table, we can see the new row is added.

```{python}
# browse the table
q = c.execute("SELECT * FROM Student").fetchall()
pd.DataFrame(q)
```