---
title: "Data Wrangling Operations in `Python`"
date: "`r Sys.Date()`"
author: Steven Wun
output:
  rmdformats::material:
    highlight: kate
  pdf_document: default
  word_document: default
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

In order to run `Python` in `R`, we need to load the `reticulate` package first.

```{r warning=FALSE}
# load the required library to run Python in R
library(reticulate)
```

We will use the same data set `airquality`, from the unit on data wrangling in `R`. We first export the data from `R` to CSV files so that the data set can be accessible for `Python`.

```{r eval=FALSE}
# write R data set to CSV file
write.table(airquality, file = "Data/airquality.csv", sep = ",", row.names = FALSE)
```



```{python}
import pandas as pd
import numpy as np
airquality = pd.read_csv("Data/airquality.csv")
airquality = airquality[['Month', 'Day', 'Ozone', 'Solar.R', 'Temp', 'Wind']]
airquality.head()
```


## Modify an R variable



## A Python chunk

This works fine and as expected. 

```{python}
x = 42 * 2
print(x) 
```

The value of `x` in the Python session is `r py$x`.
It is not the same `x` as the one in R.

## Modify a Python variable

```{python}
x = x + 18 
print(x)
```

Retrieve the value of `x` from the Python session again:

```{r}
py$x
```

Assign to a variable in the Python session from R:

```{r}
py$y = 1:5
```

See the value of `y` in the Python session:

```{python}
print(y)
```

## Python graphics

You can draw plots using the **matplotlib** package in Python.

```{python}
import matplotlib.pyplot as plt
plt.plot([0, 2, 1, 4])
plt.show()
```

```{python}
import pandas as pd
PyDf = pd.DataFrame(
  {"a": [4, 5, 6],
  "b": [7, 8, 9],
  "c": [10, 11, 12]},
  index = [1, 2, 3]
)
PyDf
```



```{python}
usvideos_url = (
"https://raw.githubusercontent.com/OrysyaStus/SeismicBlog/master/Data_Visualizations"
"/Dynamic_Visuals_Using_Date_Range_Slicers_Pt1/data/USvideos.csv"
)
usvideos = pd.read_csv(usvideos_url)
usvideos
```

```{python}
usvideos[["video_id", "title"]]
```

```{python}
url = "https://raw.githubusercontent.com/stevenkhwun/Data/main/PS4DS/four_sessions.csv"
sessions = pd.read_csv(url)
sessions
```

```{python}
sp500 = "https://raw.githubusercontent.com/stevenkhwun/Data/main/PS4DS/sp500_data.csv.gz"
sp500 = pd.read_csv(sp500)
sp500.count()
```



