---
title: "Data Wrangling Operations in `R`"
date: "`r Sys.Date()`"
author: Steven Wun
output:
  rmdformats::material:
    highlight: kate
  pdf_document: default
  word_document: default
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

In this page, we will focus on some `R` packages and functionality that is extremely useful during data wrangling processes. In particular, we will introduce:

*   The `*apply` family of `R` functions that allow us to apply functions to specific dimensions of matrices and arrays, and to data frames and lists
*   Reshaping data sets
*   Stacking data sets
*   Subsetting data sets
*   The verbs of the `dplyr` `R` package

# `*apply` Functions

The `*apply()` family is a collection of functions in `R` that is used to manipulate slices of data from matrices, arrays, lists and data frames in a repetitive way without explicitly writing loops. These functions apply a function to each element of the selected input data. Below, we focus on the following `*apply()` functions:

*   `apply()`: Returns a vector or array or list of values obtained by applying a function to margins of an array or matrix.
*   `lapply()`: Returns a list of the same length as the selected input data, each element of which is the result of applying the chosen function to the corresponding element of the selected input data.
*   `sapply()`: A more user-friendly version of `lapply()`. Returns a vector or matrix by default.
*   `mapply()`: This is the multivariate version of `sapply()`. It applies a function to multiple lists or vector arguments.

## `apply()`

The `apply()` function takes a data frame or matrix as input and returns an output of vector, list or array type. The `apply()` function is mainly used to avoid explicit uses of loops.

```{r}
# create a simple matrix
my_matrix <- matrix(c(1:20), nrow = 5)
my_matrix
# using 'apply()' to calculate the sum of each row
apply(X = my_matrix, MARGIN = 1, FUN = sum)
# using 'apply()' to calculate the sum of each column
apply(X = my_matrix, MARGIN = 2, FUN = sum)
```

In the above code chunk, we use the `apply()` function to calculate the sum of each row of the matrix by setting `MARGIN = 1`, calculate the sum of each column of the matrix by setting `MARGIN = 2`.
<BR>

In both examples, we applied the `sum()` function to the margins. However, we could have used any other function, even user-defined ones, as long as they can work with the inputs specified by the margin we use. For example, if we want to get basic summaries of each of the columns we can do

```{r}
# basic summaries of each of the columns
apply(my_matrix, 2, summary)
```

As we would expect from the `summary()` function applied to a vector, the third column of the output above has

```{r results='hold'}
# min. of the 3rd column
min(my_matrix[, 3])
# 1st quantile of the 3rd column
quantile(my_matrix[, 3], 0.25)
# median of the 3rd column
median(my_matrix[, 3])
# mean of the 3rd column
mean(my_matrix[, 3])
# 3rd quantile of the 3rd column
quantile(my_matrix[, 3], 0.75)
# max. of the 3rd column
max(my_matrix[, 3])
```

## `lapply()`

The function `lapply()` takes a vector or list `X` and applies the function `FUN` to each of its elements. Then, `lapply()` will output a list which is of the same length as `X`, where each element is the outcome of applying the function `FUN` on the corresponding element of `X`.

```{r}
# define 'First_name' and 'Last_name'
First_name <- c("John", "Jane", "Tim", "Michael", "Emma")
Last_name <- c("Doe", "Smith", "Williams", "Taylor", "Wilson")
# create a list from the names
Name_list <- list(First_name, Last_name)
Name_list
```

We use `lapply()` function to convert the names to lower case:

```{r}
# convert to lower case
Names_lower <- lapply(X = Name_list, FUN = tolower)
Names_lower
str(Names_lower)
```

In the above code chunk, we worked with a list of strings containing the first and last name of five people. We used `lapply()` together with the `tolower()` function to convert the names to lower case.

## `sapply()`

The function `sapply()` works in the same way as `lapply()` but simplifies (hence the `s`) the output to return a vector or a matrix.

```{r}
# create a simple function to raise the input to a chosen power
raise_power <- function(x, power) {
  x^power
}
```

The followings are examples of using the function.

```{r results='hold'}
# Examples
raise_power(x = 2, power = 3)
raise_power(x = 4, power = 2)
raise_power(x = 5, power = 4)
```

```{r}
# create a simple list containing numbers
numbers <- list(1:5, 6:10)
numbers
# using the 'sapply()' function to raise each element of the numbers vector to the power of 3
sapply(X = numbers, FUN = raise_power, power = 3)
```

In the above example, we created a function called `raise_power()` that takes the arguments `x` and `power` and returns `x` raised to `power`.
<br>

We then created a list called `numbers`, which contains two vectors, and used `sapply()` to apply the `raise_power()` function to each element of `numbers`. We set `power = 3` just after we have specified `X = numbers` and `FUN = raise_power`. The output is a $(5 \times 2)$ matrix, where each column represents the elements of the first and second vectors, respectively, raised to the power of three.

## `mapply()`

The function `mapply()` is the multivariate version of `sapply()`. It applies a function in parallel over a set of arguments.

```{r results='hold'}
# short demo of rep()
rep(1, 3)
rep(2, 3)
rep(3, 3)
```

```{r}
# create a 3x3 matrix
mat <- matrix(c(rep(1,3), rep(2,3), rep(3,3)), 3, 3)
mat
```

```{r}
# do the same thing via 'mapply()'
mat1 <- mapply(rep, 1:3, 3)
mat1
```

In this example, we created a matrix `mat`, using the `rep()` function thrice. We then create the same matrix `mat1` by using `mapply()`. Basically, `mapply()` applies the `rep()` function to a vector containing the numbers 1 to 3 with additional argument 3 specifying how many times each number should be repeated. Another example is

```{r}
# repeating 1 time, 4 times and 9 times
mapply(rep, 1:3, (1:3)^2)
```

## `sweep()` and `aggregate()`

The `sweep()` and `aggregate()` functions are closely related to the `apply()` family. See their help files and the useful links and resources for more information.


# Reshaping Data Sets

`R` provides a variety of methods for reshaping your data before analysis. Base `R` provides the `reshape()` function which, as the help file states reshapes a data frame between “wide” format with repeated measurements in separate columns of the same record, and “long” format with the repeated measurements in separate records. The `reshape2` `R` package provides similar functionality but with a simplified interface. To install the `reshape2` package, type in `install.packages("reshape2")`.

The `reshape2` package makes it easy to transform data between wide and long formats:

*   Wide-format data has a column for each variable
*   Long-format data has a column for possible variable types and a column for the values for those variables. However, long-format data isn’t necessarily only two columns

As you may have already figured out, you may need wide-format data for some type of analyses and long-format data for others. For example, it is easy to work with `ggplot2` with long-format data. Also, most statistical modelling functions like `lm()`, `glm()` and `gam()` work well with long-format data. However, many people often find wide-format data easier and more intuitive for recording data. Therefore, it is important to be able to work with both and be able to transform data from and into each of these two formats.

`reshape2` has two key functions:

*   `melt()`: Convert wide- to long-format data
*   `*cast()`: Convert long- to wide-format data

## Wide to Long

To illustrate how to use the `melt()` function in the `reshape2` package, we will consider the built-in `R` dataset `airquality`. This data set consists of daily air quality measurements in New York, from May to September 1973. This data is in wide format because for each data point, we have a column for each variable. In `reshape2`, to convert wide-format data to long-format data, we use the `melt()` function:

```{r warning=FALSE}
# load the 'reshape2' library
library("reshape2")
```

```{r}
# print out the first few records of the 'airquality' data frame
head(airquality)
```

```{r}
# use the 'melt' function to convert the data into long format
airquality_long <- melt(airquality)
```

```{r}
# print out the first few records of the new 'airquality_long' data frame
head(airquality_long)
```

```{r}
# print out the last few records of the new 'airquality_long' data frame
tail(airquality_long)
```

We see now our data is in long-format where we have one column named variable which records the name of the variable and a column named value which records the value of that variable. By default, the `melt()` function assumes that all columns are numeric with numeric values are variables with values. Often, we may want to know specific values and keep them as columns. We can identify these by using the `id.vars` argument. Here we can specify the `ID` variables, which are variables that identify individual rows of the data. Suppose we want to have the month and the day as ID variables:

```{r}
# use the 'melt' function to convert the data into long format with specifying ID variables Month and Day
airquality_long <- melt(airquality, id.vars = c('Month', 'Day'))
head(airquality_long)
```

We can see that by passing in the `Month` and `Day` variables into the `id.vars` argument, we kept `Month` and `Day` as columns and the rest of the data was converted into long format. We may also wish to control the column names in our long format by passing in the column names for the variable and value into the `variable.name` and `value.name` arguments respectively:

```{r}
# use the 'melt' function to convert the data into long format with specifying ID variables Month and Day
airquality_long <- melt(airquality, id.vars = c('Month', 'Day'),
                        variable.name = 'climate_var',
                        value.name = 'climate_value')
head(airquality_long)
tail(airquality_long)
```

## Long to Wide

Going from long to wide format can take a bit more thought than going from wide- to long-format data. In `reshape2`, there are multiple `cast` functions. To work with data frames, we use the `dcast()` function — there are also the `acast()` function to return a vector, matrix or array. To illustrate how to use the `dcast()` function, we work with the `airquality_long` data frame we created earlier.

The `dcast()` function uses a formula to describe the shape of the data. In this function we need to tell `dcast()` what are `ID` variables and what is the variable column that describes the measured variables. `dcast()` will use the last remaining column as the column that contains the values by default, but we can also pass this into the `value.var` argument. We can print this out and see that we can recover the original `airquality` data frame (but with the columns in different order):

```{r}
# use the 'dcast' function to convert the data back into wide format
airquality_wide <- dcast(airquality_long,
                         formula = Month + Day ~ climate_var,
                         value.var = 'climate_value')
head(airquality_wide)
```

A common problem users may face is when you cast a data set where there is more than one value per data cell. For instance, consider only using `Month` as an `ID` variable:

```{r}
# use the 'dcast' function to convert the data back into wide format
airquality_wide <- dcast(airquality_long,
                         formula = Month ~ climate_var,
                         value.var = 'climate_value')
head(airquality_wide)
```

Here we got the warning: “Aggregation function missing: defaulting to length.” By only using `Month` as the `ID` variable, we can see here that the cells are now filled with the number of data rows for each month. This is because we did not pass an aggregation function into the `fun.aggregate` argument which tells R how to aggregate the data.

If you cast your data and there are multiple values per cell, you will also need to tell `dcast()` how to aggregate the data. Depending on the aim of the analysis, we can use a range of scalar summaries, like the `mean()`, `median()`, `sum()`, etc. Here, we take the mean using the `mean()` function and we also use `na.rm = TRUE` to remove the `NA` values:

```{r}
# use the 'dcast' function to convert the data back into wide format
airquality_wide <- dcast(airquality_long,
                         formula = Month ~ climate_var,
                         value.var = 'climate_value',
                         fun.aggregate = mean,
                         na.rm = TRUE)
airquality_wide
```

We see here that we now obtained the mean value of the variables for each month.

## `tidyr` Package

There is also another popular package for data reshaping and more general data tidying, called `tidyr`. `tidyr` provides functionality for:

*   Pivoting data (i.e. reshaping);
*   Rectangling data (e.g. turning deeply nested lists into rectangular data);
*   Nesting and unnesting (e.g. converting grouped data to a few where each group becomes a single row with a nested data frame, and the reverse);
*   Splitting and combining character columns (e.g. pulling a single character column into multiple columns);
*   Making implicit missing values explicit and vice versa;

Useful Links and Resources, below, provides some links of the basic functionality from `tidyr`.

# Stacking

The `stack()` and `unstack()` functions in `R` are handy when working with data frames. Applying `stack()` to a data frame simply stacks the columns vectors on top of each other to form a single vector along with a factor indicating where each observation came from. As expected, the `unstack()` function does the complete opposite.
<br>

Below is a demonstration using the built-in `PlantGrowth` dataset. The dataset contains the weights of 30 plants from 3 different groups, namely `ctrl` (control), `trt1` (treatment 1) and `trt2` (treatment 2). Because of the way `stack()` works, we first convert the group variable from `factor` to `character` to avoid getting a warning message.

```{r}
# use the built-in 'PlantGrowth' dataset
head(PlantGrowth)
# converting group variable from factor to character
PlantGrowth$group <- as.character(PlantGrowth$group)
# stacking
stack(PlantGrowth)
# unstacking
unstack(PlantGrowth)
```

Check the help file of `stack()` for more options and examples.

# Subsetting

`R` has some powerful indexing features that allows you to quickly access object elements. In this subsection, we will learn how to efficiently select or exclude particular elements in vectors or data frames.

## Subsetting Vectors

Consider a simple numeric vector:

```{r eval=FALSE}
x <- c(1.4, 5.6, 7.8, 2.6)
```

The three ways in which we can subset elements of a vector in `R` are:

*   Using positive integers: We use a vector of positive integers to specify the index of the elements we want to return / keep.
*   Using negative integers: We use a vector of negative integers to specify the index of the elements we want to exclude.
*   Using logical vectors: We use a vector of logical values where elements are selected if the corresponding logical value is `TRUE`.

```{r}
# define x
x <- c(1.4, 5.6, 7.8, 2.6)
x
# select the 2nd and 4th element
x[c(2,4)]
# exclude the 2nd and 4th element
x[-c(2,4)]
# select the 2nd and 4th element
x[c(FALSE, TRUE, FALSE, TRUE)]
# select the elements that are strictly greater than 5
x[x > 5]
```

By running the above code, we can see that a very useful way to subset vectors is by using logical vectors. In the last line, we first evaluated the conditional statement `x > 5` and then we used the result to subset `x`.

## Subsetting Data Frames

The most basic way of subsetting a data frame in R is using square brackets `[ ]`:

```{r eval=FALSE}
dataframe[x, y]
```

where `dataframe` is a data frame in `R`, `x` is the rows that we want to be returned, and `y` is a vector of the columns that we want returned.

```{r}
# create dataframe
dataframe <- data.frame(v1 = 1:5, v2 = 6:10, v3 = 11:15)
dataframe
# select the first 3 rows
dataframe[1:3, ]
# exclude the first 3 rows
dataframe[-(1:3), ]
# select the first 3 rows and first 2 columns
dataframe[1:3, 1:2]
# select the rows using a logical vector (condition: if the variable v3 is divisible by 3)
dataframe[dataframe$v3 %% 3 == 0, ]
# select columns v1 and v3
dataframe[, c("v1", "v3")]
# select columns v1 and v3 (alternative)
dataframe[c("v1", "v3")]
```

As we have seen in “Data Structures in R,” we can also subset the rows of a data frame using the `subset()` function.

# `dplyr` Verbs

One of the most useful packages to manipulate data is `dplyr`. This package contains the so-called `dplyr` verbs:

*   `mutate()`: Adds new variables that are functions of existing variables
*   `select()`: Picks variables based on their names.
*   `filter()`: Picks cases based on their values.
*   `summarize()`: Reduces multiple values down to a single summary.
*   `arrange()`: Changes the ordering of the rows.

When working with `dplyr` verbs, we typically want to use the result of a verb applied to a data frame as input for another verb. In such cases, it is more intuitive to use the pipe `%>%` operator from the [`magrittr`](https://cran.r-project.org/web/packages/magrittr/index.html) `R` package to chain operations. The `magrittr` package is loaded automatically when `dplyr` is loaded.

The variable to the left of `%>%` operator is passed as the first argument to the function on the right of the `%>%` operator.

We have already encountered some of the those verbs and functionality when working with databases in `R`. Here, we take a more in-depth view of `dplyr`’s functionality.

Throughout this section, we use the `mtcars` data set to illustrate the capabilities of the `dplyr` verbs. Here is a preview of the dataset and its structure:

```{r}
# preview the 'mtcars' data set
head(mtcars)
str(mtcars)
```

## `mutate()`

The variable `mpg` stands for miles per gallon. Suppose we want to add another variable called `kpg` which stands for kilometres per gallon. We know that 1 mile is approximately equal to 1.61 kilometres. We can use `mutate()` to add the variable `kpg`:

```{r message=FALSE}
# loading the 'dplyr package
library(dplyr)
```

```{r}
# adding 'kpg'
mtcars1 <- mutate(mtcars, kpg = 1.61 * mpg)
head(mtcars1)
# alternative approach using the pipe '%>%' operator
mtcars2 <- mtcars %>% mutate(kpg = 1.61 * mpg)
head(mtcars2)
# mtcars1 is identical to mtcars2
identical(mtcars1, mtcars2)
```

Note that the shortcut for the pipe operator `%>%` is pressing `Shift` `Ctrl` and `M` together.

## `select()`

Suppose we only want to select the `mpg`, `cyl`, `hp`, `vs` and `gear` columns. We can use the `select()` verb to do so.

```{r}
# select columns
mtcars %>% 
  select(mpg, cyl, hp, vs, gear)
```

## `filter()`

We can use the `filter()` function to get only observations that match a condition (or multiple conditions). `filter()` is the `dplyr` equivalent to `subset()`, and allows us to write more intuitive code.

```{r}
# get observations where 'gear' is equal to 4 and 'disp' is less than 80
mtcars %>% filter(gear == 4, disp < 80)
# this is equivalent to
subset(mtcars, gear == 4 & disp < 80)
```

## `arrange()`

We can arrange our observations in any convenient way using the `arrange()` function:

```{r}
# arrange 'mtcars' in ascending order of 'mpg'
mtcars %>% 
  arrange(mpg) %>% 
  head()
# arrange 'mtcars' in descending order of 'disp'
mtcars %>% 
  arrange(desc(disp)) %>% 
  head()
```

## `summarize()`

We can compute several summaries using the `summarize()` function. For example, if we want to calculate the maximum `mpg` value, the minimum `disp` value, and the mean of `qsec` in `mtcars` we do:

```{r}
# compute summaries
mtcars %>% 
  summarize(maximum_mpg = max(mpg),
            minimum_disp = min(disp),
            average_qsec = mean(qsec))
```

`summarize()` combines well with the `group_by()` function. For example,

```{r}
# 'summarize()' and 'group_by()' functions
iris %>% 
  group_by(Species) %>% 
  summarize(Avg_Sepal_Length = mean(Sepal.Length),
            Avg_Sepal_Width = mean(Sepal.Width),
            Avg_Petal_Length = mean(Petal.Length),
            Avg_Petal_Width = mean(Petal.Width))
```

## `*_join()`

`dplyr` also provides verbs for joining data sets. Below, we use the `band_members` and `band_instruments` `tibbles` (which are `data.frame` structures) to illustrate.

```{r}
# view the data
band_members
band_instruments
```

Suppose that we want to find all rows of `band_members` for which there are matching values in `band_instruments`, and return all columns from the two `tibbles`. We can achieve that using `inner_join`:

```{r}
# inner join
band_members %>% inner_join(band_instruments)
```

We can also use `left_join()` to return all rows of `band_members` and all columns from the two tibbles, with `NA` in the new columns if there are no matches in the rows of `band_members` and `band_instruments`:

```{r}
# left join
band_members %>% left_join(band_instruments)
```

We can use `right_join()` to return all rows of `band_instruments` and all columns from the two `tibbles`, with `NA` in the new columns if there are no matches in the rows of `band_instruments` and `band_members`

```{r}
# right join
band_members %>% right_join(band_instruments)
```

Similarly, we can return all rows and columns of both `band_members` and `band_instruments` with `NA` in the new columns if there are no matches in the rows of `band_instruments` and `band_members`

```{r}
# full join
band_members %>% full_join(band_instruments)
```

Notice that the same join (with different ordering of rows) can be obtained by

```{r}
# full join
band_instruments %>% full_join(band_members)
```

All examples above also work with `data.frames`.

# Useful Links and Resources

* [Datacamp’s tutorial on the `*apply()` family and associated functions](https://www.datacamp.com/tutorial/r-tutorial-apply-family)
* [MarinStatsLectures in Youtube on the `apply` function](https://www.youtube.com/watch?v=7sJ8r6Lb7-o)
* [Xianjun Dong’s post on `reshape2` and `tidy2`](https://rpubs.com/sterding/Reshape2_and_Tidyr#:%7E:text=Reshape2%20is%20an%20R%20package,the%20general%20aggregation%20(reshape).)
* [Overview of the `tidyr` `R` package](https://tidyr.tidyverse.org/)
* [The chapter on subsetting in *Advanced R*](https://adv-r.hadley.nz/subsetting.html)
* [`dplyr`’s join pages](https://dplyr.tidyverse.org/reference/mutate-joins.html) for verbs to join `data.frames` and `tibbles`

